Grammar-Based Concept Alignment for Domain-Specific Machine Translation

### 2021-05-16 paper/paper.tex
Aarne's edits done

\begin{document}
\maketitle
\begin{abstract}
In grammar-based domain-specific MT, constructing high-quality translation lexica is a crucial but time consuming task requiring significant linguistic knowledge. 
With parallel example sentences available, statistical alignment tools can help automate part of the process, but they are not suitable for small datasets and limited when it comes to complex multiword expressions. 
In addition, the correspondences between word forms obtained in this way cannot be used directly.
Addressing these problems, we propose a grammar-based approach to this task and put it to the test in a simple translation pipeline.
\end{abstract}

\section{Introduction}
% why grammar-based approaches in DSMT
Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{ranta-etal-2020-abstract}.  
What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is often more important than coverage, they can provide strong guarantees in terms of grammatical correctness.

% importance of lexica in DSMT
Nevertheless, lexical exactness is, in this context, just as important as grammaticality. 
An important part of the design of the Controlled Natural Language (CNL) one such system's grammar describes then becomes the creation of a high-quality translation lexicon.

% how it is done right now and issues
This is often done manually, resulting in a time consuming task requiring significant linguistic knowledge. 
When the grammar is designed based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}. 
None of them is, however, suitable for the common case in which only a limited amount of example data is available.

@@

% additional advantage 1: multiwords
A further advantage of performing syntactic analysis is that we do not have to choose \textit{a priori} whether to focus on the word or phrase level. 
Instead, we can simultaneously operate at different levels of abstraction, extracting both single- and multiword, even non-contiguous, correspondences. 
For this reason, we refer to the task our system attempts to automate as \textit{Concept Alignment} (CA).

% additional advantage 2: morphology
Finally, performing morphological analysis allows us to generate translation lexica complete with grammatical category and inflection information instead of simple correspondences between word forms.

% roadmap
This paper is structured as follows. 
Section \ref{methodology} starts by giving an overview of our approach to CA and comparing it to related work to then focus on our algorithm for extracting correspondences.
After that, Section \ref{evaluation} presents the results obtained in a first evaluation of the system.
Finally, Section \ref{conclusions} summarizes our conclusions and discusses some ideas for future work.

@@

% CE
The basic use case for CA, which we refer to specifically as \textit{Concept Extraction} (CE), is the generation of a translation lexicon from a multilingual parallel text. This can be directly compared to the numerous existing word and phrase alignment techniques.

% CP
An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
While our system does implement basic CP functionalities, however, in this paper we focus on CE and restrict its application to bilingual corpora.

% contextualizing our proposal
As stated in the Introduction, most existing extraction solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the formalisms used to define the grammars for the various languages and from the lack of robustness of existing parsers.
This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Grammatical Framework (GF) \cite{ranta-etal-2020-abstract} and Universal Dependencies (UD) \cite{ws-2019-universal}.

% GF
GF is a constituency grammar formalisms and programming language in which grammars are represented as pairs of an \textit{abstract syntax}, playing the role of an interlingua, and a set of \textit{concrete syntaxes} capturing the specificities of the various natural languages. 
In the case of translation, similarly to what happens in programming language compilation, strings in the source language are \textit{parsed} to Abstract Syntax Trees (ASTs), which are then \textit{linearized} to target language strings.

% UD
UD, on the other hand, is a dependency grammar formalism meant for cross-linguistically consistent grammatical annotation.
As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each word is put in relation with the one it refers to, called its \textit{head}, via a directed labelled link specifying the syntactic relation between them.
Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word, but also on its Part-Of-Speech (POS) tag, lemma and morphological features.

@@

\subsection{Extracting concepts} 
% introducing CE algorithm
The core part of the system outlined above is of course the alignment module. 
In this section, we present our method for extracting alignments from parallel bilingual UD treebanks.
Our basic algorithm is sketched in th pseudocode below.

@@

% input
Here, the input consists of a list of priority-sorted \textit{alignment criteria}, i.e. rules to determine whether two dependency trees should be aligned with each other which will be discussed in greater detail in Section \ref{criteria}, and a pair of sentences to align represented as UD trees.
From an implementation point of view, UD trees are, rose trees where each node represents a token, obtained from the CoNLL-U files produced by the UD parser.

@@

% algorithm description
As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match any of the alignment criteria. 
If this is the case, they are added to a collection of alignments, which are represented themselves as a pair of UD (sub)trees associated with some metadata, such as the id of the sentence they were extracted from.
Such collection is what the function will return after aligning all the dependency subtrees.
The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to a lack of matching criteria.
Subtrees are sorted based on their dependency label, so to give higher priority to pairs whose heads have the same label.

% head alignment
A simple but useful refinement is that, depending on which alignment criteria a pair of trees candidate for alignment matches, the heads of the two trees may or may not also be added to the collection of alignments. 
This is important not to miss one-word correspondences that cannot be captured in any other way, for instance between the root verbs of two full sentences.
A relevant implementative detail is that, in this context, the head of a tree is not simply defined as its root. Instead, if the root is part of a compound written as two or more separate words or a verb with auxiliaries, the root nodes of the corresponding subtrees are also considered part of it.

% iterations
When working on multiple sentences, this algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be reused when working on the following ones and to keep track of the number of occurrences of each alignment throughout the entire text.

\subsubsection{Alignment criteria} \label{criteria}
% why list criteria 
While alignment criteria are customizable, to allow for a better understanding of the extraction algorithm described in the above, we describe the ones that the current implementation utilizes by default.

\paragraph{Matching UD labels}
The most obvious, but also most effective idea is to determine alignability based on comparing the dependency labels of the candidate UD tree pair. 
In particular, according to this idea, two subtrees in \textit{matching context}, i.e. attached to aligned heads, constitute an alignment if their roots share the same dependency label, meaning that they are in the same syntactic relation with their heads.
Note that, since the root of UD trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also makes it so that full sentences are always considered to align with each other, something which is desirable since we assume the parallel texts that are fed to our program to be sentence-aligned.

\paragraph{Part-Of-Speech equivalence}
% actual POS-equiv par
As noted in the above, CoNNL-U files also provide information on the grammatical categories of each word, represented as Universal POS tags \cite{petrov-etal-2012-universal}. 
Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not. 
This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all other function words, which tend to behave differently across different languages.
A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.

@@

% examples
For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
An ubiquitous example of is that of adverbs rendered as obliques, such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
\textit{Structural divergences}, where a core argument of a sentence in the source language are rendered as obliques in the source language (as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.

@@

% motivation
At a first glance, this might seem a criterion with no practical applications. 
However, it is particularly useful when, instead of starting with an empty set of correspondences, we initialize the program with some alignments that are either inserted manually or, most interestingly, obtained with some other alignment technique. 
For instance, in this way it is possible to combine the system proposed in this paper in combination with a statistical tool and give more credit to the correspondences identified by both systems.

\subsubsection{Pattern matching}
% context
So far, we described how CE can be used in a setting where the objective is to generate a comprehensive translation lexicon based on set of example sentence pairs.
We pointed out that the program can be configured so to prioritize precision or coverage, but never restricted our search to a particular type of alignments.
Nonetheless, there can be cases in which only certain syntactic structures are of interest: for instance, we might be looking for adverbs or noun phrases exclusively.

% gf-ud patterns
To handle these situations, the CE module can filter the results based on a \texttt{gf-ud} pattern.
\texttt{gf-ud} supports in fact both simple pattern matching, which is integrated in the CE module itself, and pattern replacement\footnote{documentation is available at \url{github.com/GrammaticalFramework/gf-ud}}.
Combining them, for instance by pruning the UD trees extracted by the alignment module, allows us to extract even correspondences that, given the structure of UD trees, cannot be identified by CE alone.

% predicates
To give a practically useful example, a possibility is to extract verb phrases by looking for full clauses and pruning the subtrees corresponding to subjects. 
It is also possible to perform some actual replacement to obtain \textit{predication patterns}, i.e. correspondences that specify the argument structure of verbs, such as the following English-Italian one:

@@

\subsection{Generating grammar rules}
The CE module described so far outputs pairs of UD trees. 
While converting them to GF ASTs is one of \texttt{gf-ud}'s core functionalities, building a compilable GF translation lexicon is necessary to turn such trees into grammar rules.

In order to do that, our grammar generation module requires a morphological dictionary of the languages at hand and an \textit{extraction grammar}. 
What we refer to as an extraction grammar is of course a GF grammar, whose objective is to define the set of basic categories and syntactic rules the entries of the automatically generated lexicon are going to use.
\todo{not sure how important this section is and what (if anything) to add to it}

\section{Evaluation} \label{evaluation}
% mini-roadmap
In this section, we evaluate the system proposed in the above. 
We first discuss the data used for such evaluation. 
After that, we describe our experiments, aimed at putting both the CE module \textit{per se} and the entire system from parsing to lexicon generation to the test, and present our results.

\subsection{Data} \label{data}
% PUD
Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out on treebanks instead of raw text. 
To this purpose, we use a 100-sentence subset of the Parallel UD (PUD) corpus, a set of 1000 manually annotated or validated sentences in CoNLL-U format.
Of the over 20 languages PUD treebanks are available in, we selected English, Italian and Swedish. 
Using this data limits the amount of alignment errors that are due annotation issues to a minimum, even though a small number of inconsistencies and imprecisions is present even in this corpus.

@@

\subsection{Evaluating CE} \label{ceval}
% why evaluate CE independently
While we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific tranlation, is not the only context in which CA can be put to use. 
For instance, it is easy to imagine using it to improve the UI of online translation memories.
For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the other stages stages of our lexicon generation pipeline.

% the 2 experiments
We first assess the correctness of the alignments the CE module is able to extract from the PUD treebanks, comparing our results with those obtained with a standard statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}. 
In addition, we try to quantify the impact automatic parsing has on the performance of the CE module by comparing the above results with those obtained on the DMI and CSE corpora.

% metrics
While precision and recall are two well-known performance metrics, the lack of a gold standard for CE imposes us to approximate them with the number of correct alignments the program is able to extract and the ratio between such number and the total number of alignments extracted.
However, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon and alignments that are useful for comparing the sentences but should not be used for MT. 
As an extreme example of a pair-specific alignment, consider the sentences $\langle$\textit{He missed the boat, Ha perso il treno}$\rangle$. 
In English and Italian, the idea of missing a chance is expressed using two idiomatic expression very similar to each other. 
However, the vehicle mentioned in the Italian translation is not a boat but a train (``\textit{treno}'').

@@

% details
Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files excepts with regards to tokenization. 
On the other hand, the relatively large size of the PUD treebanks makes it possible to also train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison. 
In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured to only extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.

@@

% recall
Results are less encouraging in terms of recall: the number of alignments extracted from the course plans corpora is similar to that obtained from the PUD treebank sample, despite the difference in size. 
This is explained in part by the presence, in the course plans corpora, of a large amount of very short sentences, and in part by the fact that parse errors introduced by UDPipe make it impossible to align many subsentences without a significant loss in terms of precision.

@@

Due to the small size of the datasets and the consequent low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell rather than trying to parse arbitrary sentences other course plans. 
In order to do that, we make use of GF's random AST generation functionality but at the same time manually select semantically plausible sentences to facilitate the task of the human translators.
The results of this process are two small testing corpora, one for the DMI and one for the CSE corpus, each consisting of 50 English sentences.

@@

These synthetic figures are useful to give an idea of the general quality of the translations: overall, although with relatively low scores, English-to-Swedish translation works significantly better than English-to-Italian. 
Looking back at the results reported in Section \ref{raw}, the reason for this is not immediately clear, as the difference in precision between the two language pairs is negligible on the course plans corpora.

@@

Furthermore, a problem with using the BLEU score as the only evaluation metric is the fact that it makes no distinction between content and function words, thus not allowing an evaluation focused specifically on the extracted concepts. 
The small size of the corpus, however, allows for some error analysis. 
From the participants's observations about the kind of errors encountered when manually editing the automatic translation, summarizes in Table \ref{obs}, we can conclude that while most errors are in fact due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors.
This explains the significant difference observed in the cumulative BLEU scores shown in Table \ref{tableu}.

@@

\section{Conclusions} \label{conclusions}
We presented a syntax-based alignment method with a focus on its applications in domain-specific translation lexica generation.

Compared to the existing statistical tools, our system has the following advantages:
\begin{itemize} \setlength\itemsep{0.1em}
  \item it performs consistently well even on extremely small parallel corpora
  \item it is able to simultaneously extract correspondences between individual words, multiword expressions and full phrases, even when their constituents are discontinuous
  \item in conjunction with \texttt{gf-ud} pattern matching, it can be used to extract specific types of correspondences, such as predication patterns  
  \item it can automatically generate compilable, morphology-aware GF translation lexica
  \item it can be configured to easily handle systematic, possibly language-pair or corpus specific translation divergences.
\end{itemize}

@@

An obvious possible development is to improve the current implementation of CP, optimizing it for its different use cases: propagating alignments to a new language looking for correspondences using a translation of the same text they were extracted from or using a different text in the same domain.
An alternative to the latter is to make CE, now working on bilingual texts, $n$-lingual.

When large amounts of data ar available, using our system in conjunction with a statistical tool seems promising. 
As discussed in Section \ref{ka}, this is already partially supported and it could prove useful to develop CA as an actual hybrid system.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Grammar-based domain-specific MT systems are a common use case for CNL.
High-quality translation lexica are a crucial part of such systems, but involve time consuming work and significant linguistic knowledge. 
With parallel example sentences available, statistical alignment tools
can help automate part of the process, but they are not suitable for
small datasets and do not always perform well with complex multiword expressions. 
In addition, the correspondences between word forms obtained in this way cannot be used directly.
Addressing these problems, we propose a grammar-based approach to this task and put it to test in a simple translation pipeline.
\end{abstract}

\section{Introduction}
% why grammar-based approaches in DSMT
Grammar-based translation pipelines such as those based on Grammatical Framework (GF) have been successfully employed in domain-specific Machine Translation (MT) \cite{ranta-etal-2020-abstract}.  
What makes these systems well suited to the task is the fact that, when we constrain ourselves to a specific domain, where precision is often more important than coverage, they can provide strong guarantees of correctness.

% importance of lexica in DSMT
However, lexical exactness is, in this context, just as important as grammaticality. 
An important part of the design of a Controlled Natural Language (CNL) is the creation of a high-quality translation lexicon, preserves both semantics and grammatical correctness.

% how it is done right now and issues
A translation lexicon is often built manually, which is a time consuming task requiring significant linguistic knowledge. 
When the task is based on a corpus of parallel example sentences, it is possible to automate part of this process by means of statistical word and phrase alignment techniques \cite{brown-etal-1993-mathematics, och-ney-2000-improved, dyer-etal-2013-simple}. 
None of them is, however, suitable for the common case in which only a small amount of example data is available --- typically, which just one occurrence of each relevant lexical item.

@@

% additional advantage 1: multiwords
A further advantage of performing syntactic analysis is that we do not have to choose \textit{a priori} whether to focus on the word or phrase level. 
Instead, we can simultaneously operate at different levels of abstraction, extracting both single- and multiword, even non-contiguous, correspondences. 
For this reason, we refer to the task our system attempts to automate as \textit{Concept Alignment} (CA). 
A \textit{concept} is a semantic unit expressed by a word or a construction, which is also a unit of \textit{compositional translation}, where translation is performed by mapping concepts to concepts in a shared syntactic structure.

% additional advantage 2: morphology
Concepts are to be conceived as \textit{lemmas} equipped with morphological variations, instead of fixed word forms or phrases.
This  allows us to generate translation lexica complete with grammatical category and inflection so that correct target language forms can be selected in each syntactic context.

% roadmap
This paper is structured as follows. 
Section \ref{methodology} starts by giving an overview of our approach to CA and comparing it with related work, followed by a description of our CA algorithm.
Section \ref{evaluation} presents the results obtained in an evaluation of the system.
Section \ref{conclusions} summarizes our conclusions and discusses some ideas for future work.

@@

% CE
The basic use case for CA, which we refer to specifically as \textit{Concept Extraction} (CE), is the generation of a translation lexicon from a multilingual parallel text.
This is analogous to the well-known earlier word and phrase alignment techniques.

% CP
An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
While our system does implement basic CP functionalities, in this paper we focus on CE and restrict its application to bilingual corpora.

% contextualizing our proposal
As stated in the Introduction, most existing CE solutions are based on statistical approaches and are, as a consequence, unsuitable for small datasets. 
Grammar-based approaches, making use of parallel treebanks and collectively referred to as \textit{tree-to-tree alignment methods}, have also been proposed \cite{tiedemann2011bitext}, but have historically suffered from the inconsistencies between the formalisms used to define the grammars of different languages and from the lack of robustness of parsers.
This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Grammatical Framework (GF) \cite{ranta-2011} and Universal Dependencies (UD) \cite{ws-2019-universal}.

% GF
GF is a constituency grammar formalism and programming language in which grammars are represented as pairs of an \textit{abstract syntax}, playing the role of an interlingua, and a set of \textit{concrete syntaxes} capturing the specificities of the various natural languages. 
In the case of translation, similarly to what happens in programming language compilation, strings in the source language are \textit{parsed} to Abstract Syntax Trees (ASTs), which are then \textit{linearized} to target language strings.

% UD
UD, on the other hand, is a dependency grammar formalism meant for cross-linguistically consistent grammatical annotation.
As opposed to constituency, \textit{dependency} is a word-to-word correspondence: each word is put in relation with the one it depends on, called its \textit{head}, via a directed labelled link specifying the syntactic relation between them.
Importantly for our application, the standard format for UD trees, CoNNL-U, gives information not only on the syntactic role of each word, but also on its Part-Of-Speech (POS), lemma, and morphological features.

@@

\subsection{Extracting concepts} 
% introducing CE algorithm
The core part of the system outlined above is the alignment module. 
Its function is to extract alignments from parallel bilingual UD treebanks.
The outline of the algorithm is given in the following pseudocode:

@@

% input
Here, the input consists of a list of priority-sorted \textit{alignment criteria}, i.e. rules to determine whether two dependency trees should be aligned with each other, and a pair $(t,u)$ of UD trees to align.
An example alignment criterion is sameness of syntactic label, for instance, that subjects are aligned with subjects and objects with objects; the details will be discussed in Section \ref{criteria},  
From an implementation point of view, UD trees are, rose trees (trees with arbitrary numbers of branches) where each node represents a word with its dependents as subtrees.
The rose tree is is easily obtained from the CoNLL-U notation that UD parsers produce.

@@

% algorithm description
As a first step, the program checks whether the two full sentence trees can be aligned with each other, i.e. if they match the alignment criteria. 
If this is the case, they are added to a collection of alignments, which are represented as pairs of UD (sub)trees associated with some metadata, such as the id of the sentence they were extracted from.
Such a collection is what the function will return after aligning all the dependency subtrees.
The same procedure is applied recursively to all pairs of immediate subtrees of each sentence, until the leaves are reached or alignment is no longer possible due to lack of matching criteria.
Subtrees are sorted based on their dependency label, to give higher priority to pairs whose heads have the same label.

% head alignment
A simple but useful refinement is that, depending on which alignment criteria a pair of trees matches, the heads of the two trees may or may not also be added to the collection of alignments. 
This is done in order not to miss one-word correspondences that cannot be captured in any other way, for instance between the root verbs of two full sentences.
A relevant implementation detail is that, in this context, the head of a tree is not simply defined as its root. Instead, if the root is part of a compound written as two or more separate words or a verb with auxiliaries, the root nodes of the corresponding subtrees are also considered parts of it.

% iterations
When working on multiple sentences, the algorithm can be applied in an iterative fashion, so that knowledge gathered when a sentence pair is aligned can be used when working on later sentences and also to keep track of the number of occurrences of each alignment throughout the entire text.

\subsubsection{Alignment criteria} \label{criteria}
% why list criteria 
While the alignment criteria are customizable, to allow for a better understanding of the extraction algorithm described above, we explain the criteria that our implementation utilizes by default.

\paragraph{Matching UD labels}
The most obvious, but also most effective idea is to determine alignability based on comparing the dependency labels of the candidate UD tree pair. 
In particular, according to this idea, two subtrees in \textit{matching context}, i.e. attached to aligned heads, constitute an alignment if their roots share the same dependency label, meaning that they are in the same syntactic relation with their heads.
Note that, since the root of UD trees is always attached to a a fake node with an arc labelled \texttt{root}, this criterion also implies that full sentences are always considered to align with each other.
This is desirable since we assume that the parallel texts that are fed to our program are sentence-aligned.

\paragraph{Part-Of-Speech equivalence}
% actual POS-equiv par
As noted above, the CoNNL-U notation provides information on the grammatical categories of each word, represented as Universal POS tags \cite{petrov-etal-2012-universal}. 
Intuitively, if the nodes of two trees in matching contexts have the same POS tags, the two trees are more likely to correspond to each other than if not. 
This is especially true if we focus, for instance, solely on the open class words (defined as in the UD documentation \footnote{\url{universaldependencies.org/u/pos/all.html}}), thus ignoring prepositions, determiners, auxiliary verbs and all function words, which tend to behave differently across different languages.
A useful relation to define between dependency trees is, then, that of \textit{POS-equivalence}: two dependency trees $t$, $u$ are POS-equivalent if $M_1 = M_2 \neq \emptyset$, where $M_i$ is defined as the multiset of POS tags of all the open class word nodes of $t_i$.

@@

% examples
For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
A ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) are translated prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
\textit{Structural divergences}, where a direct object in one language is rendered as an oblique in the other, as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.

@@

% motivation
At a first glance, this might look like a criterion with no practical applications. 
However, it is particularly useful when, instead of starting with an empty set of correspondences, we initialize the program with some alignments that are either inserted manually or, most interestingly, obtained with some other alignment technique. 
For instance, in this way it is possible to combine the system proposed in this paper with a statistical tool and give more credit to correspondences identified by both systems.

\subsubsection{Pattern matching}
% context
So far, we have described how CE can be used in a setting where the objective is to generate a comprehensive translation lexicon based on set of example sentence pairs.
We pointed out that the program can be configured to prioritize precision or coverage, but we did not restrict our search to a particular type of alignments.
However, there are cases in which only certain syntactic structures are of interest: for instance, we might be looking for adverbs or noun phrases exclusively.

% gf-ud patterns
To handle such cases, the CE module can filter the results based on a \texttt{gf-ud} \textit{tree patterns}.
\texttt{gf-ud} supports in fact both simple pattern matching, which is integrated in the CE module itself, and pattern replacement\footnote{documentation is available at \url{github.com/GrammaticalFramework/gf-ud}}.
Combining them, for instance by pruning the UD trees extracted by the alignment module, allows us to extract correspondences that cannot be identified by CE alone.

% predicates
For example, pattern matching can extract verb phrases by looking for full clauses and dropping the subtrees corresponding to subjects. 
By means of replacements, one can obtain \textit{predication patterns}, that is, correspondences that specify the argument structure of verbs, such as the following English-Italian one:

@@

\subsection{Generating grammar rules}
The CE module described so far outputs pairs of UD trees. 
While converting them to GF ASTs is one of \texttt{gf-ud}'s core functionalities, building a compilable GF translation lexicon is needed to turn such trees into grammar rules.

In order to do that, our grammar generation module uses a morphological dictionary of the languages at hand and an \textit{extraction grammar}. 
The extraction grammar is a GF grammar that defines the syntactic categories and functions the entries of the automatically generated lexicon are build with.
For example, entries can be prepositional phrases (PP) or verb phrases (VP) constructed by the following GF functions:
\begin{verbatim}
  PrepNP : Prep -> NP -> PP 
         # case head
  PrepPP : VP   -> PP -> VP 
         # head obl
\end{verbatim}
The dependency labels appended to the function types instruct \texttt{gf-ud} to build GF trees from UD trees that match these labels.

\section{Evaluation} \label{evaluation}
% mini-roadmap
In this section, we evaluate the system proposed above. 
We first discuss the data used in the evaluation. 
After that, we describe our experiments, aimed at putting both the CE module \textit{per se} and the entire system from parsing to lexicon generation to the test, and present our results.

\subsection{Data} \label{data}
% PUD
Because we want part of our evaluation to be independent from the quality of UD parsing, some of the experiments are carried out on treebanks instead of raw text. 
To this purpose, we use a 100-sentence subset of the Parallel UD (PUD) corpus, a set of 1000 manually annotated or validated sentences in CoNLL-U format.
Of the over 20 languages PUD treebanks are available in, we selected English, Italian and Swedish. 
Using this data limits the amount of alignment errors that are due to annotation issues to a minimum, even though a small number of inconsistencies and errors is present even in this corpus.

@@

\subsection{Evaluating CE} \label{ceval}
% why evaluate CE independently
While we focus mostly on the MT applications of CA, automatic translation, and much less GF-based domain-specific translation, is not the only context in which CA can be put to use. 
For instance, it is easy to imagine using it to build translation memories used by human translators.
For this reason, a first set of experiments is aimed at evaluating the alignments obtained with our CA module independently from the other stages of our lexicon generation pipeline.

% the 2 experiments
We first assess the correctness of the alignments the CE module is able to extract from the PUD treebanks, comparing our results with those obtained with a statistical tool, \texttt{fast\_align} \cite{dyer-etal-2013-simple}. 
In addition, we try to quantify the impact of automated UD parsing on the performance of the CE module by comparing the above results with those obtained on the DMI and CSE corpora.

% metrics
While precision and recall are two well-known performance metrics, the lack of a gold standard for CE forces us to assess the correctness of alignments manually before calculating the ratio between correctly extracted alignments and the total of extracted alignments.
What is more, since some alignments are only correct in the specific context of the sentence pair in which they occur, we make a further distinction between correct alignments that are relevant for a translation lexicon and alignments that are useful for comparing the sentences but should not be used for MT. 
As an extreme example of a pair-specific alignment, consider the sentences $\langle$\textit{He missed the boat, Ha perso il treno}$\rangle$. 
In English and Italian, the idea of missing a chance is expressed using two idiomatic expression very similar to each other. 
However, the vehicle mentioned in the Italian translation is not a boat but a train (``\textit{treno}'').

@@

% details
Of course, \texttt{fast\_align} does not make use of the information present in the CoNLL-U files except with regards to tokenization. 
On the other hand, the relatively large size of the PUD treebanks makes it possible also to train the statistical tool on the full dataset instead of just using the chosen 100-sentences subset, allowing for a fairer comparison. 
In both cases, \texttt{fast\_align} is run with the recommended parameters and the CE program is configured only to extract one-to-many and many-to-one word alignments, given that \texttt{fast\_align} does not align larger phrases.

@@

% recall
The results are less encouraging in terms of recall: the number of alignments extracted from the course plan corpora is similar to that obtained from the PUD treebank sample, despite the difference in size. 
This is explained in part by the presence, in the course plans corpora, of a large amount of very short sentences, and in part by the fact that parse errors introduced by UDPipe make it impossible to align many subsentences without a significant loss in terms of precision.

@@

Due to the small size of the datasets and the consequently low coverage of the extracted lexicon, we generate the sentences to translate directly in the GF shell rather than trying to parse arbitrary sentences other course plans. 
In order to do that, we make use of GF's random AST generation functionality but at the same time manually select semantically plausible sentences to facilitate the task of the human translators.
The results of this process are two small testing corpora, one for the DMI and one for the CSE corpus, each consisting of 50 English sentences.

@@

These synthetic figures are useful to give an idea of the general quality of the translations: overall, although with relatively low scores, English-to-Swedish translation works significantly better than English-to-Italian. 
Looking back at the results reported in Section \ref{raw}, the reason for this is not immediately clear, as the difference in precision between the two language pairs is negligible in the course plan corpora.

@@

Furthermore, a problem with using the BLEU score as the only evaluation metric is the fact that it makes no distinction between content and function words, thus not allowing an evaluation focused specifically on the extracted concepts. 
The small size of the corpus, however, allows for some error analysis. 
From the participants' observations about the kind of errors encountered when manually editing the automatic translation, summarizes in Table \ref{obs}, we can conclude that while most errors are in fact due to wrong alignments, the main difference between two corpora lies in the number of translations that only contain grammatical errors.
This explains the significant difference observed in the cumulative BLEU scores shown in Table \ref{tableu}.

@@

\section{Conclusions} \label{conclusions}
We have presented a syntax-based alignment method with a focus on its applications in domain-specific translation lexic generation.

Compared with the existing statistical tools, our system has the following advantages:
\begin{itemize} \setlength\itemsep{0.1em}
  \item it performs consistently well even on extremely small parallel corpora
  \item it is able to simultaneously extract correspondences between individual words, multiword expressions and full phrases, including discontinuous constituents
  \item in conjunction with \texttt{gf-ud} pattern matching, it can be used to extract specific types of correspondences, such as predication patterns  
  \item it can automatically generate compilable, morphology-aware GF translation lexica
  \item it can be configured to easily handle systematic, possibly language-pair or corpus specific translation divergences.
\end{itemize}

@@

An obvious possible development is to improve the current implementation of concept propagation (CP), optimizing it for its different use cases: propagating alignments to a new language looking for correspondences using a translation of the same text they were extracted from or using a different text in the same domain.
An alternative to the latter is to make CE, now working on bilingual texts, $n$-lingual.

When large enough amounts of data ar available, using our system in conjunction with a statistical tool seems promising. 
As discussed in Section \ref{ka}, this is already partially supported and it could prove useful to develop CA as an actual hybrid system.
